State of the Art in Concurrent Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-17 paper/paper.tex
SprayList section, some benchmarking

\section{Lock-free Priority Queues}

@@

@@

\begin{itemize}
\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
      However, as the original code was written for the MIPS architecture, we chose to
      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
      instead.
\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
      According to the authors, the commercial Noble library includes a more efficient
      version of this data structure that can also handle duplicate priority values.
\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
      SkipList implementation. The aim of this implementation is to minimize contention in
      calls to \lstinline|DeleteMin|.
\end{itemize}

In each test run, the examined priority queue was initially filled with $2^{15}$
elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
for a total of $10$ seconds, where all \lstinline|Insert|
operations within this context implicitly choose a key uniformly at random from
the range of all 32-bit integers. Each run was repeated for a total of $10$ times %TODO: Verify this.
and we report on the average throughput.

All benchmarks were compiled with \verb|-O3| on GCC 4.8.2. Evaluations took place
on an 80-core Intel Xeon machine.

>>>>

\section{Lock-free Priority Queues} \label{sec:lockfree}

@@

The SprayList is another recent approach towards a relaxed priority queue by \citeauthor{alistarhspraylist}
\cite{alistarhspraylist}. Instead of the distributed approach described in the previous section,
the SprayList is based on a central data structure, and uses a random walk in \lstinline|DeleteMin|
in order to spread accesses over the $O(P \log^3 P)$ first elements with high probability, where $P$
is again the number of participating threads.

\citeauthor{fraser2004practical}'s lock-free SkipList \cite{fraser2004practical} again serves as the
basis for the priority queue implementation. The full source code is available online\footnote{
\url{https://github.com/jkopinsky/SprayList}}. In the SprayList, \lstinline|Insert| calls are simply
forwarded to the underlying SkipList.

The \lstinline|DeleteMin| operation however executes a random walk, also called a \emph{spray}, the
purpose of which is to spread accesses over a certain section of the SkipList uniformly such that
collisions between multiple concurrent \lstinline|DeleteMin| calls are unlikely. This is achieved by
starting at some initial height, walking a randomized number of steps, descending a randomized number of levels,
and repeating this procedure until a node $n$ is reached on the lowest level. If $n$ is not deleted,
it is logically deleted and returned. Otherwise, a \emph{spray} is either reattempted, or the thread
becomes a cleaner, traversing the lowest level of the SkipList and physically removing logically deleted
nodes it comes across. A number of dummy nodes are added to the beginning of the list in order to counteract
the algorithm's bias against initial items.

The \emph{spray} parameters are chosen such that with high probability, one of the $O(P \log^3 P)$
first elements is returned, and that each of these elements is chosen roughly uniformly at random.
The final effect is that accesses to the data structure are spread out, reducing contention and resulting
in a noticeably lower number of \ac{CAS} failures in comparison to strict priority queues described
in Section \ref{sec:lockfree}.

The authors do not provide any statement as to the linearizability (or other concurrent correctness
criteria) of the SprayList, and it is not completely clear how to define it since no
sequential semantics are given.

Benchmarks show promising results: the SprayList scales well at least up to 80 threads,
and performs close (within a constant factor) to an implementation using a random remove instead
of \lstinline|DeleteMin|, which the authors consider as the performance ideal.

% Maybe mention that P log^3 P is a very loose bound.

@@

\begin{itemize}
\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
      by the C++ standard library, protected by a single global lock.
\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
      However, as the original code was written for the MIPS architecture, we chose to
      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
      instead.
\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
      According to the authors, the commercial Noble library includes a more efficient
      version of this data structure that can also handle duplicate priority values.
\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
      SkipList implementation. The aim of this implementation is to minimize contention in
      calls to \lstinline|DeleteMin|.
\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
      SkipList using random walks to spread data accesses
      incurred by \lstinline|DeleteMin| calls. Code provided by \citeauthor{alistarhspraylist} is
      available on Github\footnote{\url{https://github.com/jkopinsky/SprayList}}.
\end{itemize}

In each test run, the examined priority queue was initially filled with $2^{15}$
elements. We then ran a tight loop of $50\%$ insertions and $50\%$ deletions
for a total of $10$ seconds, where all \lstinline|Insert|
operations within this context implicitly choose a key uniformly at random from
the range of all 32-bit integers. This methodology seems to be the de facto standard for concurrent % TODO: Similar phrasing as linden :(
priority queue benchmarks
\cite{alistarhspraylist,linden2013skiplist,shavit2000skiplist,sundell2003fast}.
Each run was repeated for a total of $10$ times
and we report on the average throughput.

All benchmarks were compiled with \verb|-O3| using GCC 4.8.2. Evaluations took place
on an 80-core Intel Xeon machine. The benchmarking code was adapted from \citeauthor{linden2013skiplist}'s
benchmarking suite and is available at \url{https://github.com/schuay/seminar_in_algorithms}.
